TRIK Studio: Technical Introduction

### 2017-02-10 2017-trikStudioEnglish/trikStudioEnglish.tex
Final polishing

USB communication is implemented using libusb\footnote{http://libusb.org/ (accessed: 07.02.2017)} library, Bluetooth communication is based on QextSerialPort\footnote{https://github.com/qextserialport/ (accessed: 07.02.2017)} library, Wi-Fi communication is implemented over TCP and UDP protocols using QtNetwork library, complex communication protocols (like configuration of TRIK robots before running programs) are built upon Qt State Machine Framework. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implementations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

An interpreter is translating visual diagrams into a sequence of commands for a target device (see Fig.~\ref{image:interpretersTSArch}). The hierarchy of devices is build in a way that it does not matter to an interpreter if it works with a real device of a virtual one. Each device is represented by a C++ class containing code for interaction with this device. These device classes are grouped according to robotic kits they belong to and put in plug-in modules. Device classes use communication subsystem for data interchange over physical channels (see Section~\ref{chapter:communications}) and 2D and 3D simulators API when working with virtual robot models.

@@

The interpreter of control-flow programs starts with finding initial block of the program and then iterates according to the control flow defined by diagram arrows. Currently executed block is highlighted in the editor to show users execution status. For each block visited special factories create objects that implement the behavior of this block. In general such an object searches for a ready-to-work robot device that it needs, runs appropriate commands and passes execution to one of the blocks that the current block has an outgoing arrow to (depending on the block's semantics). Within this process no difference is made if the device is a real one or a virtual one, so the same interpretation code works for execution of a program in two out of three modes (see Section~\ref{chapter:generalDescription}).

@@

The data-flow interpreter works in two steps. First, the interpreted diagram is being validated, implementation objects for diagram blocks are created (they are used to translate the diagram into a sequence of robot commands similar to as it was discussed for the control-flow programs interpreter). Implementation objects are connected to each other via Qt signals and slots in the same order as corresponding blocks are connected on a diagram (the Publisher-subscriber design pattern is used here). On the second step the interpreter executes each block that does not have any incoming data flows, and then each such block in its turn activates blocks that it has outgoing data flows to when the data tokens are ready. Blocks execution is performed in a pseudo-parallel way over a centralized message queue. This implementation detail distinguishes this language from all other industrial solutions (e.g. in Microsoft Robotics Developer Studio diagrams are deployed as a set of independent web services~\cite{jackson2007microsoft}). We chose this approach because this programming environment targets low-performance controllers of educational robots, that don't embrace true parallel execution of multiple programs at once. Nevertheless, the language has a process fork block that is useful while implementing complex control models mentioned above. This block should be used as a low-level tool for execution control.

@@

As mentioned before, a visual diagram is built from blocks and arrows, defining program's \textit{control flow graph}~\cite{steven1997advanced}. All algorithmic operators like conditions, loops, switches, etc. are defined by arrows. For instance, to create an infinite loop one should create an arrow from a block to one of the previous blocks, to create \texttt{while-do} and \texttt{do-while} loops or loops with \texttt{break} statement inside it is enough to simply connect one block inside a loop with a block from the outside. It is obvious that each arrow on a diagram corresponds to a \texttt{goto} statement in the resulting code. But the code containing \texttt{goto} statements is very hard to understand, and especially is not suitable for teaching programming basics, so we decided to avoid using them in generated code as much as possible. We cannot eliminate \texttt{goto} statements completely because it is possible to create a TRIK Studio diagram that could not be expressed using standard algorithmic operators (there are algorithms for automatic translation of \texttt{goto}-containing programs into structural programs, but the result is not much better from the teaching programming point of view). On the other hand, not all modern textual programming languages support \texttt{goto} statements (for example, JavaScript does not). We call \textit{structural} diagrams that could be expressed using structural algorithmic statements (\texttt{if-then}, \texttt{if-then-else}, \texttt{while-do loop}, \texttt{do-while loop}, \texttt{while-break loop}, \texttt{switch}). Textual code generated from them we also call \textit{structural}. Otherwise we say that the diagram (and the code generated from it) is \textit{not structural}. If structural code could not be generated from a given diagram, users are notified with a warning.

@@

\begin{itemize}
    \item Code generation subsystem should be organized in a way to minimize time and effort for adding of a new generator. Preferably is should be possible even for an external developer to do it.
    \item For each language (where applicable) the subsystem should support two generation modes: structural diagrams should be translated into structural code, not structural diagrams -- into not structural code. The success of structuring the diagram should not depend on its size and complexity.
\end{itemize}

The first task is purely of architectural nature. The way to solve it is shown on Fig.~\ref{image:generatorsArchitecture}. The main idea here is to divide code generation process into two steps. First, the diagram is translated into a representation independent of a target language --- a \textit{semantic tree}. Semantic trees reorder the control flow graph model into a tree model regardless of whether the diagram is structural or not. If it is, a parent node in the semantic tree always corresponds to an execution block in generated code, and its children nodes correspond to statements of this block, which exclude \texttt{goto}. Let's outline the code generation process shown in Fig.~\ref{image:generatorsArchitecture}.

@@

A code generator gets a diagram as input data. First, the diagram is traversed in depth-first order by a validation component (build based on the Visitor design pattern). Then, for each computable expression used inside block property values parsing and type inference procedure is performed. If the diagram or any code in any property value contains an error, users are notified and the code generation process terminates. If the diagram is syntactically correct, it is passed to a control flow analyzer, that is described below. Depending on if the diagram can be translated into structural code or not, an intermediate generator of code-independent diagram representation is selected, which extracts the semantic tree from the control flow graph. Finally, the semantic tree is printed into a target textual language. To achieve it, a template-based approach is used, so to create a new generator a programmer has to provide a set of such code templates and define a generator specification. For example, generation templates for condition expression if C looks like this:
\begin{verbatim}
    if (@@CONDITION@@) {
        @@THEN_BODY@@
    } else {
        @@ELSE_BODY@@
    }
\end{verbatim}

@@

One obvious limitation of the target textual language for code generation is that it should be an imperative one: the chain of blocks must be translated into a sequence of target language expressions, which is better achieved using sequential composition. Nevertheless, an approach with intermediate step with generation into intermediate representation could work here too. For example, to support code generation into a pure functional language one could add a continuation element into the intermediate representation and add intermediate generator from visual diagrams into continuation passing style (CPS) representation. After that code generation templates should be updated to support all necessary languages.

@@

Last important subsystem which is described in this paper is automatic checker of constraints for TRIK Studio programs. This subsystem allows to turn a world model in a simulator to an exercise with specified success and failure conditions, which can be shared among students, and solutions (in a form of visual programs) can be automatically checked against these success conditions in the simulator, thus providing feedback without an intervention from a teacher. To create an exercise one needs to specify two things:

@@

Let's briefly describe predicates, function symbols and elementary triggers used in the language. Predicate symbols are divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an ``item $x$ is located inside area $y$'' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other ``syntactic sugar''.
\end{itemize}

>>>>

USB communication is implemented using \texttt{libusb}\footnote{http://libusb.org/ (accessed: 07.02.2017)} library, Bluetooth communication is based on \texttt{QextSerialPort}\footnote{https://github.com/qextserialport/ (accessed: 07.02.2017)} library, Wi-Fi communication is implemented over TCP and UDP protocols using \texttt{QtNetwork} library, complex communication protocols (like configuration of TRIK robots before running programs) are built upon \texttt{Qt State Machine Framework}. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implementations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

An interpreter translates visual diagrams into a sequence of commands for a target device (see Fig.~\ref{image:interpretersTSArch}). The hierarchy of devices is build in a way that it does not matter to an interpreter if it works with a real device of a virtual one. Each device is represented by a C++ class containing code for interaction with this device. These device classes are grouped according to robotic kits they belong to and put in plug-in modules. Device classes use communication subsystem for data interchange over physical channels (see Section~\ref{chapter:communications}) and 2D and 3D simulators API when working with virtual robot models.

@@

The interpreter of control-flow programs starts with finding the initial block of the program and then iterates according to the control flow defined by diagram arrows. Currently executed block is highlighted in the editor to show users execution status. For each block visited special factories create objects that implement the behavior of this block. In general such an object searches for a ready-to-work robot device that it needs, runs appropriate commands and passes execution to one of the blocks that the current block has an outgoing arrow to (depending on the block's semantics). Within this process no difference is made if the device is a real one or a virtual one, so the same interpretation code works for execution of a program in two out of three modes (see Section~\ref{chapter:generalDescription}).

@@

The data-flow interpreter works in two steps. First, the interpreted diagram is being validated, implementation objects for diagram blocks are created (they are used to translate the diagram into a sequence of robot commands similar to as it was discussed for the control-flow programs interpreter). Implementation objects are connected to each other via Qt signals and slots in the same order as corresponding blocks are connected on a diagram (the Publisher-subscriber design pattern is used here). On the second step the interpreter executes each block that does not have any incoming data flows, and then each such block in its turn activates blocks that it has outgoing data flows to when the data tokens are ready. Blocks execution is performed in a pseudo-parallel way over a centralized message queue. This implementation detail distinguishes this language from all other industrial solutions (e.g. in Microsoft Robotics Developer Studio diagrams are deployed as a set of independent web services~\cite{jackson2007microsoft}). We chose this approach because this programming environment targets low-performance controllers of educational robots, that don't embrace true parallel execution of multiple programs at once. Nevertheless, the language has a process fork block that is useful while implementing complex control models mentioned above. This block is a low-level tool for execution control.

@@

As mentioned before, a visual diagram is built from blocks and arrows, defining program's \textit{control flow graph}~\cite{steven1997advanced}. All algorithmic operators like conditions, loops, switches, etc. are defined by arrows. For instance, to create an infinite loop one should create an arrow from a block to one of the previous blocks, to create \texttt{while-do} and \texttt{do-while} loops or loops with \texttt{break} statement inside it is enough to simply connect one block inside a loop with a block from the outside. It is obvious that each arrow on a diagram corresponds to a \texttt{goto} statement in the resulting code. But the code containing \texttt{goto} statements is very hard to understand, and especially is not suitable for teaching programming basics, so we decided to avoid using them in generated code as much as possible. We cannot eliminate \texttt{goto} statements completely because it is possible to create a TRIK Studio diagram that could not be expressed using standard algorithmic operators (there are algorithms for automatic translation of \texttt{goto}-containing programs into structural programs, but the result is not much better from the teaching programming point of view). On the other hand, not all modern textual programming languages support \texttt{goto} statements (for example, JavaScript does not). We call \textit{structural} diagrams that could be expressed using structural algorithmic statements (\texttt{if-then}, \texttt{if-then-else}, \texttt{while-do loop}, \texttt{do-while loop}, \texttt{while-break loop}, \texttt{switch}). Code generated from structural diagrams we also call \textit{structural}. Otherwise we say that the diagram (and the code generated from it) is \textit{not structural}. If structural code could not be generated from a given diagram, users are notified with a warning.

@@

\begin{itemize}
    \item Code generation subsystem should be organized in a way to minimize time and effort for adding of a new generator. Preferably it should be possible even for an external developer to do it.
    \item For each language (where applicable) the subsystem should support two generation modes: structural diagrams should be translated into structural code, non-structural diagrams -- into non-structural code. The success of structuring the diagram should not depend on its size and complexity.
\end{itemize}

The first task is purely of architectural nature. The solution it is shown on Fig.~\ref{image:generatorsArchitecture}. The main idea  is to divide code generation process into two steps. First, the diagram is translated into a representation independent of a target language --- a \textit{semantic tree}. Semantic trees reorder the control flow graph model into a tree model regardless of whether the diagram is structural or not. If it is, a parent node in the semantic tree always corresponds to an execution block in generated code, and its children nodes correspond to statements of this block, which exclude \texttt{goto}. An outline of the code generation process is shown in Fig.~\ref{image:generatorsArchitecture}.

@@

A code generator gets a diagram as input data. First, the diagram is traversed in a depth-first order by a validation component. Then, for each computable expression used inside block property values parsing and type inference procedure is performed. If the diagram or any code in any property value contains an error, users are notified and the code generation process terminates. If the diagram is syntactically correct, it is passed to a control flow analyzer, that is described below. Depending on if the diagram can be translated into structural code or not, an intermediate generator of code-independent diagram representation is selected, which extracts the semantic tree from the control flow graph. Finally, the semantic tree is printed into a target textual language. To achieve it, a template-based approach is used, so to create a new generator a programmer has to provide a set of such code templates and define a generator specification. For example, generation templates for condition expression in C looks like this:
\begin{verbatim}
    if (@@CONDITION@@) {
        @@THEN_BODY@@
    } else {
        @@ELSE_BODY@@
    }
\end{verbatim}

@@

@@

Last important subsystem described in this paper is automatic checker of constraints for TRIK Studio programs. This subsystem allows to turn a world model in a simulator to an exercise with specified success and failure conditions, which can be shared among students, and solutions (in a form of visual programs) can be automatically checked against these success conditions in the simulator, thus providing feedback without an intervention from a teacher. To create an exercise one needs to specify two things:

@@

Let us briefly describe predicates, function symbols and elementary triggers used in the language. Predicate symbols are divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an ``item $x$ is located inside area $y$'' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other ``syntactic sugar''.
\end{itemize}
